Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-29 ltx/data_paper/DataPaper.tex
general edits to the data paper

Gaussian distributions of the state disturbance and the observation noise results into the following conditional distributions, where the normalising factors are ignored since scaling by a constant coefficient does not alter the location of the maximum with respect to the parameters.
 p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&= \mid\boldsymbol\Sigma_{\epsilon}\mid^{-\frac{1}{2}}  \exp\left({-\frac{1}{2}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)^\top\boldsymbol\Sigma_{\epsilon}^{-1}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)}\right),\\
p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}} \nonumber \\

@@

Note that disturbance covariance matrix, $\boldsymbol\Sigma_e$, can be written as the product of the unknown parameter, $\sigma_d^2$ with the constant matrix, $\tilde{\boldsymbol\Sigma}_e$, where the constant part can be  calculated using the estimated support for disturbance covariance function, $\gamma$ from the correlation analysis (see equation \eqref{eq:EM-MMGFourier2}). By expanding the exponent and taking $2\ln$ we get

@@

2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d)&=-n_x\ln\sigma_d^2-\ln\mid\tilde{\boldsymbol\Sigma}_e\mid-\frac{1}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_{t+1}\nonumber \\
&+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta 
+\frac{2\xi}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t \nonumber \\
&-\frac{1}{\sigma_e^2}\boldsymbol\theta^\top \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta\nonumber \\
&-\frac{\xi^2}{\sigma_d^2}\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t. \label{eq:CondititionDist2}

@@

In the next step the above equation is rearranged in order to isolate the covariance and the cross-covariance terms, enabling the use of smoother outputs. Taking trace and rearranging, applying the invariant cyclic permutations property of the trace, equations \eqref{eq:CondititionDist1} and \eqref{eq:CondititionDist2} become

@@

 2 \ln &p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\sigma_d) =-n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace\mathbf x_{t+1} \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta \nonumber \\

@@

Substituting  \eqref{eq:Qfunctiontrace1} and \eqref{eq:Qfunctiontrace2} into \eqref{eq:QIntermsofJointDist} and taking the expectation of the log likelihood function for all time instants gives the lower-bound which is to be maximised in the M-step. 

@@

& -\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\

@@

The third term in \eqref{eq:Voldermont} is obtained using the same method in \cite{Shumway2000}, i.e. by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1} \mathbf{\hat{x}}_{t+1}^\top \mathbf C^\top$ to the expectation of the second term in \eqref{eq:Qfunctiontrace1} . By defining $\Xi$-variables as 

@@

equation \eqref{eq:Voldermont} can be written in a form of

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top\right] \right\rbrace\nonumber \\

@@

Covariance and cross-covariance terms in equation \eqref{eq:VoldermontwitXivariables}, consist of both linear terms, $\boldsymbol\Xi_0$, $\boldsymbol\Xi_1$ and $\boldsymbol\Xi_2$ and non-linear terms, $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and  $\boldsymbol\Xi_5$. Finding linear $\boldsymbol\Xi$-variables is straightforward as they can be computed directly using smoother outputs, i.e. $\hat{\mathbf x}_t^b $, $\mathbf P_t^b$ and $\mathbf M_t^b$ \cite{Gibsona2005}. To show how it can be done, the smoothed covariance matrix at time $t+1$ can be written as

@@

 \mathbf E_{\Theta'}\left[\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\mathbf P_{t+1}^b+\mathbf {\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\\

@@

However calculation of non-linear $\boldsymbol\Xi$-variables is not straightforward as in linear case. Therefore, an additional  Taylor expansion step to the URTSS algorithm is also created to approximate $\mathbf q(\mathbf{x}_t)$ at each time instant which is to be used in  calculations of $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and $\boldsymbol\Xi_5$. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function  around the smoothed state, $\hat{\mathbf x}_t^b$ at each time instant, i.e.

@@

@@

substituting \eqref{eq:SigmoidTaylor} in \eqref{eq:QmatrixForSigmapoints},  an approximation for $\mathbf q(\mathbf x_t)$ around the smoothed state $\hat{\mathbf{x}}_t^b$ can be obtained

@@

Having approximated $\mathbf q\left(\mathbf x_t\right)$, by substituting  \eqref{eq:qTaylor} into equations (\ref{eq:defofXi3}-\ref{eq:defofXi5}) an approximation to the non-linear $\boldsymbol\Xi$-variables can be made. The three nonlinear variables are dealt with in turn. To start with, first consider equation \eqref{eq:defofXi3}, substituting in equation \eqref{eq:qTaylor} for $\mathbf q\left(\mathbf x_t\right)$, the term under the summation can be approximated by

@@

noting that $\phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)$ is a scalar, equation \eqref{eq:Xi3Derivation1} can be rewritten as 

@@

Substituting for  $\mathbf E_{\boldsymbol\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_{t+1}^\top\right]$ from equation \eqref{eq:EM-PrimeryXi1derivation1} and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ yields

@@

To find an approximation for $\boldsymbol\Xi_4$, the same approach is applied by using \eqref{eq:qTaylor} for the nonlinear experession in equation \eqref{eq:defofXi4} giving

@@

&= \mathrm{term1} + \mathrm{term2} + \mathrm{term3} + \mathrm{term4}

@@

It should be noted that when applying expectation, term1 is constant, and term2 and term3 disapears as $\mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\right]= \mathbf 0$. However in term4 to use the covariance estimate from the smoother the $\mathbf{x}_t$ terms should be brought together. To do this equation \eqref{eq:-EMterm4} can be rewritten as 

@@

\begin{align}\label{eq:EM-Lambdat}
	{\boldsymbol \Lambda}_t^{(i)}&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t^b)d\mathbf r'.

@@

  \mathbf{E}_{\boldsymbol\Theta'}\left[ \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\right] &= \mathbf q^\top(\mathbf {\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)\nonumber \\

@@

and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ yields

@@

 \boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[\mathbf q^\top(\mathbf{\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf \mathbf q(\mathbf{\hat x}_t^b)+\boldsymbol\varLambda_t\right]\label{eq:Xi4Ap}. 

@@

Finally, to calculate $\boldsymbol\Xi_5$, the nonlinear term can be treated in a smae way, using the expression for $\mathbf q(\mathbf x_t)$ in \eqref{eq:defofXi5} and rearranging gives

@@

 \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\approx\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)+\int_\Omega \boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'

@@

taking expected value of \eqref{eq:Xi5derivation1} gives

@@

&+\int_\Omega \boldsymbol \phi^\top(\mathbf r') \mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}' 

@@

substituting for $\mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\right] $ from equation \eqref{eq:EM-PrimeryXi0derivation1} at time $t$, and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ gives

@@

By computing $\boldsymbol\Xi$-variables, calculation of the lower bound is complete. The lower bound now can be used in M-step to find a set of new parameters. 
\subsection{M-Step}
% In the M-step, an update $\boldsymbol\Theta'$ is chosen such that $\mathcal{Q}\left(\boldsymbol\Theta',\boldsymbol\Theta'\right)>\mathcal{Q}\left(\boldsymbol\Theta,\boldsymbol\Theta'\right)$, this will guarantee that the likelihood increases.
The objective of the M-step is to maximize the computed lower bound in E-step, whcich consequenly increases the log-likelihood function $p(\mathbf Y;\boldsymbol\Theta)$.
% on the joint likelihood function, $p(\mathbf X,\mathbf Y;\boldsymbol \Theta)$, where $\mathbf X$ is the set of all states, $\mathbf Y$ is the set of all observations, and $\boldsymbol \Theta$ denotes the set of all unknown parameters, i.e., connectivity kernel parameters $\boldsymbol\theta$, synaptic dynamic $\xi$, disturbance variance $\sigma^2_d$ and the observation noise variance $\sigma^2_{\epsilon}$.
                                                                                                                                                                                                                                                                                                                                                                                                                                                                            Consider first the maximisation of $\mathcal{Q}$-function with respect to $\xi$, differentiating equation \eqref{eq:VoldermontwitXivariables} with respec to the synaptic parameter gives

@@

In order to maximise $\mathcal{Q}$-function with respect to the connectivity kernel parameters, $\boldsymbol\theta$, differentiation is taken as follows

@@

noting that variable $\boldsymbol\Xi_4$ is symmetric by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields

@@

The condition under which $\boldsymbol\Xi_4$ is invertible is discussed later in this section. Now maximising the lower bound with respect to $\sigma_{\epsilon}^2$ gives

@@

and setting the above equation to zero and solving for $\sigma_{\epsilon}^2$ gives the result

@@

Finally differentiating the lower bound with respec to $\sigma_d^2$ gives

@@

setting above to zero and solving for $\sigma_d^2$ completes the maximisation step

@@

In order to confirm calculated parameters are indeed maxima of the lower bound, the second derivative test is carried out for each parameter, giving

@@

$\boldsymbol\Xi_2$ is positive definite under the persistent excitation condition of state vectors, which is guaranteed by the disturbance, $\mathbf e$, and $\tilde{\boldsymbol\Sigma}_e^{-1}$ is positive definite by construction (Note that the inverse of a positive definite matrix is also positive definite \cite{Horn2005}), therefore by \cite{Coope1994}

@@

@@

\caption{{\bf Disturbance support estimates for variyng observation noise variance.} Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf a) $\sigma_{\varepsilon}^2$=0. (\textbf b) $\sigma_{\varepsilon}^2$=0.1 . (\textbf c) $\sigma_{\varepsilon}^2$=0.2 .  (\textbf d) $\sigma_{\varepsilon}^2$=0.23 .  (\textbf e) $\sigma_{\varepsilon}^2$=0.3 . (\textbf f) $\sigma_{\varepsilon}^2$=0.4 .}

@@

\caption{{\bf Disturbance support approximation for }$\sigma_{\varepsilon}^2<$0.23. Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf{a}) Estimated support is shown by red circle, nonzero values after thresholding are shown by the black area. (\textbf{b}) True support.}

@@

\caption{{\bf Cross-section of disturbance covariance support estimation}. The normalized true and the estimated width are shown by solid and dashed lines respectively.}

@@

\begin{figure}

@@

\caption{{\bf Kernel support approximation for different values of synaptic time constant, $\tau$ and observation noise variance, $\sigma^2_{\varepsilon}$}. Each subplot shows normalised connectivity kernel estimate. (\textbf a) $\sigma_{\varepsilon}^2$=0. (\textbf b) $\sigma_{\varepsilon}^2$=0.1 . (\textbf c) $\sigma_{\varepsilon}^2$=0.18.}

>>>>

The distribution of the initial states is assumed to be Gaussian and known. Therefore, the last term on the right-hand side of \eqref{eq:EM-QwithInitialstates} can be omitted as it does not depend on the parameter set, $\mathcal Q$-function can be then rewritten as
Gaussian distributions of the field disturbance and the observation noise result into the following conditional distributions, where the constant normalising factors are dropped as multiplying by a constant does not alter the location of the maximum.
\begin{align}\label{eq:EM-twicelog1}
 p\left(\mathbf y_{t+1}|\mathbf x_{t+1};\sigma_{\epsilon}\right)&= \mid\boldsymbol\Sigma_{\epsilon}\mid^{-\frac{1}{2}}  \exp\left({-\frac{1}{2}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)^\top\boldsymbol\Sigma_{\epsilon}^{-1}\left(\mathbf y_{t+1}-\mathbf C\mathbf  x_{t+1}\right)}\right),
\begin{align}\label{eq:EM-twicelog2}
p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta , \xi, \sigma_d)&= \mid\boldsymbol\Sigma_{e}\mid^{-\frac{1}{2}} \nonumber \\

@@

Note that the disturbance covariance matrix, $\boldsymbol\Sigma_e$, can be written as the product of the unknown parameter, $\sigma_d^2$ with a constant matrix, $\tilde{\boldsymbol\Sigma}_e$, where the constant part can be calculated (\parham{from \eqref{eq:DisturbanceCovariance}}) using the estimated support for the disturbance covariance function, $\gamma$, from the correlation analysis (see \eqref{eq:EM-MMGFourier2}). This simplifies the M-step significantly as the estimation of the entire covariance matrix  reduces to the estimation of the single scalar  $\sigma_d^2$. By expanding the exponent in \eqref{eq:EM-twicelog1} and \eqref{eq:EM-twicelog2}, twice the logarithm of the results give

@@

2\ln p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta ,\xi, \sigma_d)&=-n_x\ln\sigma_d^2-\ln\mid\tilde{\boldsymbol\Sigma}_e\mid-\frac{1}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_{t+1}+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta\nonumber \\
&+\frac{2\xi}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t -\frac{1}{\sigma_e^2}\boldsymbol\theta^\top \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta-\frac{2\xi}{\sigma_d^2} \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\boldsymbol\theta\nonumber\nonumber \\
& -\frac{\xi^2}{\sigma_d^2}\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t. \label{eq:CondititionDist2}
% &-\franc{\xi^2}{\sigma_d^2}\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf x_t. \label{eq:CondititionDist2}

@@

In the next step the above equations are rearranged to isolate the covariance and the cross-covariance terms, allowing the smoother outputs to be used. In order to permute the matrices, the scalar nature of the likelihood is exploited, this way the trace operator can be applied to express the likelihood in a more compact form. Taking trace and rearranging, applying the invariant cyclic permutations property of the trace, equations \eqref{eq:CondititionDist1} and \eqref{eq:CondititionDist2} become

@@

 2 \ln &p(\mathbf x_{t+1}|\mathbf x_{t};\boldsymbol \theta, \xi, \sigma_d) =-n_x\ln\sigma_d^2-\frac{1}{\sigma_d^2}\mathrm{tr}\left\lbrace\mathbf x_{t+1} \mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace+\frac{2}{\sigma_d^2}\mathbf x_{t+1}^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q( \mathbf x_t)\boldsymbol\theta \nonumber \\

@@

Substituting  \eqref{eq:Qfunctiontrace1} and \eqref{eq:Qfunctiontrace2} back into \eqref{eq:QIntermsofJointDist} and taking the expectation of the log-likelihood function at all time instants gives the required lower-bound. Note that the expectation distributes over the trace sum. 

@@

& -\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}^b) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}^b)^\top+\mathbf C \mathbf P_{t+1}^b\mathbf C^\top\right] \right\rbrace\nonumber \\

@@

The third term in \eqref{eq:Voldermont} is obtained using the same method in \cite{Shumway2000}, i.e. by adding and subtracting $\mathbf C \mathbf{\hat{x}}_{t+1}^b \mathbf{\hat{x}}_{t+1}^{b\top} \mathbf C^\top$ to the expectation of the second term in \eqref{eq:Qfunctiontrace1} and then exploiting the expression given in \eqref{eq:EM-Xi0derivation1} . By defining $\Xi$-variables as 

@@

equation \eqref{eq:Voldermont} can be written in a more compact form

@@

 \mathcal Q(\boldsymbol \Theta,\boldsymbol\Theta')&=-\frac{1}{\sigma_{\epsilon}^2}\mathrm{tr}\left\lbrace\boldsymbol\sum_{t=0}^{T-1}\left[ (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}^b) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}^b)^\top+\mathbf C \mathbf P_{t+1}^b\mathbf C^\top\right] \right\rbrace\nonumber \\

@@

The covariance and cross-covariance terms in \eqref{eq:VoldermontwitXivariables}, consist of both the linear terms, $\boldsymbol\Xi_0$, $\boldsymbol\Xi_1$ and $\boldsymbol\Xi_2$ and non-linear terms, $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and  $\boldsymbol\Xi_5$. Finding the linear $\boldsymbol\Xi$-variables is straightforward as they can be computed directly from the smoother outputs, i.e. $\hat{\mathbf x}_t^b $, $\mathbf P_t^b$ and $\mathbf M_t^b$ \cite{Gibsona2005}. To show how it can be done, the smoothed covariance matrix at time $t+1$ can be written as

@@

 \mathbf E_{\Theta'}\left[\mathbf x_{t+1}\mathbf x_{t+1}^\top\right]=\mathbf P_{t+1}^b+\mathbf {\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}

@@

However calculation of the non-linear $\boldsymbol\Xi$-variables is not straightforward as in the linear case. Therefore, an additional Taylor expansion step to the URTSS algorithm is also created to linearise $\mathbf q(\mathbf{x}_t)$, enabling the  calculations of $\boldsymbol\Xi_3$, $\boldsymbol\Xi_4$ and $\boldsymbol\Xi_5$ using the smoother outputs. This is accomplished by taking the first-order truncated Taylor-series expansion of the sigmoidal activation function at each time instant  around the smoothed state, $\hat{\mathbf x}_t^b$, i.e.

@@

% \begin{align}\label{eq:SigmoidDerivative}
%  f'(\boldsymbol \phi^\top\left(\mathbf r'\right)&\hat{\mathbf x}_t^b)= \varsigma f(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b)\left( 1-f(\boldsymbol \phi^\top\left(\mathbf r'\right)\hat{\mathbf x}_t^b\right).
% \end{align}

@@

substituting \eqref{eq:SigmoidTaylor} in \eqref{eq:EM-Stateequationnonlinearity},  an approximation for $\mathbf q(\mathbf x_t)$ around the smoothed state $\hat{\mathbf{x}}_t^b$ can be obtained

@@

Having approximated $\mathbf q\left(\mathbf x_t\right)$, by substituting  \eqref{eq:qTaylor} into equations (\ref{eq:defofXi3}-\ref{eq:defofXi5}) an approximation to each of the non-linear $\boldsymbol\Xi$-variable can be made. The three non-linear variables are dealt with in turn. To start, first consider \eqref{eq:defofXi3}, substituting in \eqref{eq:qTaylor} for $\mathbf q\left(\mathbf x_t\right)$, the term under the summation sign can be approximated by

@@

noting that $\phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)$ is scalar, \eqref{eq:Xi3Derivation1} can be rearranged in the form 

@@

Substituting for  $\mathbf E_{\boldsymbol\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_{t+1}^\top\right]$ from \eqref{eq:EM-PrimeryXi1derivation1} and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ yields

@@

To find an approximation for $\boldsymbol\Xi_4$, a similar approach is used where the non-linear expression in \eqref{eq:defofXi4} is replaced by \eqref{eq:qTaylor}  giving

@@

&= \mathrm{term1} + \mathrm{term2} + \mathrm{term3} + \mathrm{term4},

@@

It should be noted that when applying expectation, term1 is constant, and term2 and term3 vanish as $\mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\right]= \mathbf 0$. However in term4 to use the covariance estimate from the smoother the $\mathbf{x}_t$ terms should be permuted. To do this \eqref{eq:-EMterm4} can be rewritten as 

@@

	{\boldsymbol \Lambda}_t^{(i)}&=\int_{\Omega} \boldsymbol \Psi(\mathbf r')\phi_i(\mathbf r')f'(\boldsymbol\phi^\top(\mathbf r')\mathbf {\hat x}_t^b)d\mathbf r' \label{eq:lambdatildei}.

@@

  \mathbf{E}_{\boldsymbol\Theta'}\left[ \mathbf q^\top(\mathbf x_t)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\right] &\approx \mathbf q^\top(\mathbf {\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)\nonumber \\

@@

summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ yields

@@

 \boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[\mathbf q^\top(\mathbf{\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf{\hat x}_t^b)+\boldsymbol\varLambda_t\right]\label{eq:Xi4Ap}. 

@@

Finally, to calculate $\boldsymbol\Xi_5$, the non-linear term can be treated in a similar way, using the expression for $\mathbf q(\mathbf x_t)$ in \eqref{eq:defofXi5} and rearranging gives

@@

 \mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf x_t)\approx\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf {\hat x}_t^b)+\int_\Omega \boldsymbol \phi^\top(\mathbf r') (\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'.

@@

Taking expected value of \eqref{eq:Xi5derivation1} yields

@@

&+\int_\Omega \boldsymbol \phi^\top(\mathbf r') \mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\right]\tilde{\boldsymbol\Sigma}_e^{-1}\boldsymbol{\Psi}(\mathbf{r}')f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'.

@@

Substituting for $\mathbf E_{\Theta'}\left[(\mathbf x_t - \mathbf  {\hat x}_t^b)\mathbf x_t^\top\right] $ from \eqref{eq:EM-PrimeryXi0derivation1} at time $t$, and summing over $t \in \left\lbrace 0, \dots, T-1\right\rbrace $ gives

@@

By computing $\boldsymbol\Xi$-variables, the approximation of the lower bound is complete. The lower bound now can be used in the M-step to find a set of new parameters. 
\subsubsection{M-Step}
The objective of the M-step is to maximise the computed lower bound in the E-step, which consequently increases the log-likelihood function $\ln p(\mathbf Y;\boldsymbol\Theta)$.                                                                                                                                                                                                                                                                                                                                                                                                                                  Consider first the maximisation of the $\mathcal{Q}$-function with respect to $\xi$, differentiating \eqref{eq:VoldermontwitXivariables} with respect to the synaptic parameter gives

@@

In order to maximise the $\mathcal{Q}$-function with respect to the connectivity kernel parameters, $\boldsymbol\theta$, differentiation is taken as follows

@@

noting that the variable $\boldsymbol\Xi_4$ is symmetric by construction, equating \eqref{eq:Qpartialtheta} to the zero vector yields

@@

The condition under which $\boldsymbol\Xi_4$ is invertible will be discussed later in this section (see below \eqref{eq:EM-traceinequality}). Now maximising the lower bound with respect to $\sigma_{\epsilon}^2$ gives

@@

and setting the above equation to zero and solving for $\sigma_{\epsilon}^2$ yields

@@

Finally differentiating the lower bound with respect to $\sigma_d^2$ gives

@@

setting the above to zero and solving for $\sigma_d^2$ completes the maximisation step.

@@

In order to confirm that the calculated solutions maximise the lower bound, a second derivative test is performed for each of the parameter giving

@@

$\boldsymbol\Xi_2$ is positive definite under the persistent excitation condition of the state vectors, which is guaranteed by the disturbance, $\mathbf{e}_t(\mathbf r)$, and $\tilde{\boldsymbol\Sigma}_e^{-1}$ is positive definite by construction (Note that the inverse of a positive definite matrix is also positive definite \cite{Horn2005}), therefore by \cite{Coope1994}
\begin{equation}\label{eq:EM-traceinequality}

@@

For initialisation a bounded random state sequence is employed to ensure that the initial parameter estimates describe a stable kernel, although the quality of such initial estimate is poor, it provides an initialisation that typically ensues a satisfactory convergence of the algorithm, as illustrated numerically in \S\ref{sec:EM-Results}. The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. Here the change in parameter estimates, $L^2$-norm in the case of the connectivity kernel weights, is used to halt the execution of the algorithm. Besides, the lower bound on the likelihood function is also monitored to further investigate the convergence of the EM algorithm.

An approximate for the computational complexity of the algorithm is given in Table~\ref{table:ComputationalComplexity}.  In this table only the computational complexity of the variables specific to the presented algorithm are shown, and the computational complexity of the URTSS can be found in \cite{Sarkka2010}. Note that the computational complexity of the non-linear terms depend on the number of points, $n_r$, used for the spatial discretisation. Calculation of the non-linear terms is clearly the slowest part of the algorithm where $n_x$ is the critical dimension, more specifically the matrix $\boldsymbol\Xi_4 $ has the complexity that grows with the order of four in $n_x$. Although the number of states  is decoupled from the number of observation locations (recording electrodes in this case) , and the number of basis functions for the connectivity kernel decomposition, it  grows quadratically in two dimensional space for the equally spaced basis functions.
%  Increasing the joint likelihood function increases the likelihood function $p(\mathbf Y;\boldsymbol \Theta)$ \cite{McLachlan1997}. 
 \begin {table}
\scalebox{1}{\begin{tabular}{lll}
\hline \hline
Variable &Equation&Order\\ \hline\\
$\mathbf q(\mathbf{x}_t)$&\eqref{eq:EM-Stateequationnonlinearity}&$O(n_xn_{\theta}n_r)$\\
$\boldsymbol\Xi_0, \boldsymbol\Xi_1, \boldsymbol\Xi_2$&\eqref{eq:Xi0Gibson}, \eqref{eq:Xi1Gibson}, \eqref{eq:Xi2Gibson}&$O(Tn_x^2)$\\
$\boldsymbol\Xi_3, \boldsymbol\Xi_5$&\eqref{eq:Xi3Ap},\eqref{eq:Xi5Ap} &$O(Tn_x^3)+O(Tn_x^2n_r)+O(Tn_xn_{\theta}n_r)$ \\
$\boldsymbol\Xi_4$&\eqref{eq:Xi4Ap}&$O(Tn_x^4n_{\theta})+O(Tn_x^3n_{\theta}^2)+O(Tn_x^3n_{\theta}n_r)$\\
$\hat{\xi}$&\eqref{eq:xihat}&$O(n_x^3+n_{\theta}^3)$\\
$\hat{\boldsymbol\theta} $&\eqref{eq:thetahat}&$O(n_{\theta}^3)$\\
$\hat{\sigma}_{\epsilon}^2$&\eqref{eq:sigmaepsilonhat}&$O(Tn_x^2n_y)+O(Tn_xn_y^2)$\\
$\hat{\sigma}_{d}^2$&\eqref{eq:sigmadhat}&$O(n_x^3)+O(n_{\theta}^2)$\\
\hline \hline
\end{tabular}}
 \caption {Computational complexity of the EM algorithm} 
\label{table:ComputationalComplexity}
\end {table}

@@

\caption{{\bf Disturbance support estimates for different values of observation noise variance.} Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf a) $\sigma_{\varepsilon}^2$=0. (\textbf b) $\sigma_{\varepsilon}^2$=0.1 . (\textbf c) $\sigma_{\varepsilon}^2$=0.2 .  (\textbf d) $\sigma_{\varepsilon}^2$=0.23 .  (\textbf e) $\sigma_{\varepsilon}^2$=0.3 . (\textbf f) $\sigma_{\varepsilon}^2$=0.4 .}

@@

\caption{{\bf Disturbance support approximation for }$\sigma_{\varepsilon}^2<$0.23. Each subplot shows the support of the disturbance covariance function convolved twice with the sensor kernel. (\textbf{a}) Estimated support is shown by red circle, non-zero values after thresholding are shown by the black area. (\textbf{b}) True support.}

@@

\caption{{\bf Cross-section of normalised disturbance covariance function estimation}. The actual and the estimated covariance functions (obtained by numerical deconvolution via \eqref{eq:EM-MMGFourier2}) are shown by solid and dashed lines respectively.}

@@

@@

\caption{{\bf Kernel support approximation for different values of the synaptic time constant, $\tau$ and the observation noise variance, $\sigma^2_{\varepsilon}$}. Each subplot shows the normalised connectivity kernel estimate. (\textbf a) $\sigma_{\varepsilon}^2$=0. (\textbf b) $\sigma_{\varepsilon}^2$=0.1 . (\textbf c) $\sigma_{\varepsilon}^2$=0.18.}
